MADA 8060E Project

### 2023-04-07 code/analysis_code/analysis.qmd
Submit Part 4

```{r, echo=FALSE, message=FALSE}
# load R packages
library(here) 
library(readxl)
library(ggplot2)
library(skimr) 
library(plotly)
library(gt)
library(knitr)
library(dplyr)
library(kableExtra)
library(gridExtra)
library(ggpubr)
library(janitor)
```

@@

>>>>

```{r, echo=FALSE, message=FALSE}
# load R packages
library(here) 
library(readxl)
library(ggplot2)
library(skimr) 
library(plotly)
library(gt)
library(knitr)
library(dplyr)
library(kableExtra)
library(gridExtra)
library(ggpubr)
library(janitor)
library(tidymodels)
library(dotwhisker)
library(tidyverse)
library(caret)
```

@@

# Analyze State Race Data with Google Trend HIV Search

```{r}
state_race_gtrend<-readRDS(here("data", "processed_data", "state_race_gtrend.rds"))
gtrends_2021_2022 <- readxl::read_excel(here("data","raw_data", "Misc", "Gtrends_2021_2022.xlsx")) 
p_2022 <- gtrends_2021_2022 %>% filter(Year %in% c(2022))

linear_reg() %>% 
  set_engine("keras")
lm_mod <- linear_reg()

# predict 2022 black-white
lm_fit <- 
  lm_mod %>% 
  fit(black_white_rateratio ~ Gtrend + Year + State , data = state_race_gtrend)
lm_fit
tidy(lm_fit)
tidy(lm_fit) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))

mean_pred <- predict(lm_fit, new_data = p_2022)
conf_int_pred <- predict(lm_fit, new_data = p_2022, type = "conf_int")
glance(lm_fit) %>%
  dplyr::select(adj.r.squared, sigma, AIC, BIC, p.value)

plot_data <- 
  p_2022 %>% 
  bind_cols(mean_pred) %>% 
  bind_cols(conf_int_pred)

compare <- state_race_gtrend  %>% filter(Year %in% c(2020)) %>% select(State, black_white_rateratio, hispanic_white_rateratio) 
plot_data_c <- merge(plot_data,compare,by=c("State"),all=T)

south<-plot_data_c %>% filter(State %in% c('Alabama','Arkansas','Delaware','District of Columbia','Florida','Georgia',
                                         'Kentucky','Louisiana','Maryland','Mississippi','North Carolina','Oklahoma',
                                         'South Carolina','Tennessee','Texas','Virginia','West Virginia'))
northeast<-plot_data_c %>% filter(State %in% c('Connecticut', 'Maine', 'Massachusetts', 'New Hampshire', 'New Jersey', 'New York', 
                                             'Pennsylvania', 'Rhode Island', 'Vermont'))
west<-plot_data_c %>% filter(State %in% c('Alaska', 'Arizona', 'California', 'Colorado', 'Hawaii', 'Idaho', 'Montana', 'Nevada', 'New Mexico',
                                        'Oregon', 'Utah', 'Washington', 'Wyoming'))
midwest<-plot_data_c %>% filter(State %in% c('Illinois', 'Indiana', 'Iowa', 'Kansas', 'Michigan', 'Minnesota', 'Missouri', 'Nebraska', 'North Dakota',
                                           'Ohio','South Dakota', 'Wisconsin'))

south22<-ggplot(south, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Black-White Rate-Ratio") + 
  scale_x_discrete(labels=c('AL','AR','DE','DC','FL','GA', 'KY','LA','MD','MS','NC','OK',
                                         'SC','TN','TX','VA','WV')) +
  ggtitle(label="SOUTH: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = black_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))

northeast22<-ggplot(northeast, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Black-White Rate-Ratio") + 
  scale_x_discrete(labels=c('CT', 'ME', 'MA', 'NH', 'NJ', 'NY', 'PA', 'RI', 'VT')) +
  ggtitle(label="NORTHEAST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = black_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))

west22<-ggplot(west, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Black-White Rate-Ratio") + 
  scale_x_discrete(labels=c('AK', 'AZ', 'CA', 'CO', 'HI', 'ID', 'MT', 'NV', 'NM', 
                            'OR', 'UT', 'WA', 'WY')) +
  ggtitle(label="WEST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = black_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))

midwest22<-ggplot(midwest, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Black-White Rate-Ratio") + 
  scale_x_discrete(labels=c('IL', 'IN', 'IA', 'KS', 'MI', 'MN', 'MO', 'NE', 'ND',
                            'OH','SD', 'WI')) +
  ggtitle(label="MIDWEST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = black_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Black-White Rate Ratio, 2020'))

predict_2022_blackwhite<-ggarrange(northeast22, south22, west22, midwest22, ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")
predict_2022_blackwhite
figure_file = here("results","Predict2022_BWratio.png")
ggsave(filename = figure_file, plot=predict_2022_blackwhite)

# predict 2022 hispanic-white
lm_fit <- 
  lm_mod %>% 
  fit(hispanic_white_rateratio ~ Gtrend + Year + State , data = state_race_gtrend)
lm_fit
tidy(lm_fit)
tidy(lm_fit) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2)) 
mean_pred <- predict(lm_fit, new_data = p_2022)
conf_int_pred <- predict(lm_fit, new_data = p_2022, type = "conf_int")
glance(lm_fit) %>%
  dplyr::select(adj.r.squared, sigma, AIC, BIC, p.value)

plot_data <- 
  p_2022 %>% 
  bind_cols(mean_pred) %>% 
  bind_cols(conf_int_pred)

compare <- state_race_gtrend  %>% filter(Year %in% c(2020)) %>% select(State, black_white_rateratio, hispanic_white_rateratio) 
plot_data_c <- merge(plot_data,compare,by=c("State"),all=T)

south<-plot_data_c %>% filter(State %in% c('Alabama','Arkansas','Delaware','District of Columbia','Florida','Georgia',
                                         'Kentucky','Louisiana','Maryland','Mississippi','North Carolina','Oklahoma',
                                         'South Carolina','Tennessee','Texas','Virginia','West Virginia'))
northeast<-plot_data_c %>% filter(State %in% c('Connecticut', 'Maine', 'Massachusetts', 'New Hampshire', 'New Jersey', 'New York', 
                                             'Pennsylvania', 'Rhode Island', 'Vermont'))
west<-plot_data_c %>% filter(State %in% c('Alaska', 'Arizona', 'California', 'Colorado', 'Hawaii', 'Idaho', 'Montana', 'Nevada', 'New Mexico',
                                        'Oregon', 'Utah', 'Washington', 'Wyoming'))
midwest<-plot_data_c %>% filter(State %in% c('Illinois', 'Indiana', 'Iowa', 'Kansas', 'Michigan', 'Minnesota', 'Missouri', 'Nebraska', 'North Dakota',
                                           'Ohio','South Dakota', 'Wisconsin'))

south22<-ggplot(south, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Hispanic-White Rate-Ratio") + 
  scale_x_discrete(labels=c('AL','AR','DE','DC','FL','GA', 'KY','LA','MD','MS','NC','OK',
                                         'SC','TN','TX','VA','WV')) +
  ggtitle(label="SOUTH: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))

northeast22<-ggplot(northeast, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Hispanic-White Rate-Ratio") + 
  scale_x_discrete(labels=c('CT', 'ME', 'MA', 'NH', 'NJ', 'NY', 'PA', 'RI', 'VT')) +
  ggtitle(label="NORTHEAST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))

west22<-ggplot(west, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Hispanic-White Rate-Ratio") + 
  scale_x_discrete(labels=c('AK', 'AZ', 'CA', 'CO', 'HI', 'ID', 'MT', 'NV', 'NM', 
                            'OR', 'UT', 'WA', 'WY')) +
  ggtitle(label="WEST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))

midwest22<-ggplot(midwest, aes(x = State)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "Hispanic-White Rate-Ratio") + 
  scale_x_discrete(labels=c('IL', 'IN', 'IA', 'KS', 'MI', 'MN', 'MO', 'NE', 'ND',
                            'OH','SD', 'WI')) +
  ggtitle(label="MIDWEST: Predict 2022") + 
  theme_light() + 
  geom_point(aes(y = hispanic_white_rateratio, color="red")) + 
  theme(legend.title = element_blank()) + 
  scale_colour_discrete(labels=c('Actual Hispanic-White Rate Ratio, 2020'))

predict_2022_hispanicwhite<-ggarrange(northeast22, south22, west22, midwest22, ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")
predict_2022_hispanicwhite
figure_file = here("results","Predict2022_HWratio.png")
ggsave(filename = figure_file, plot=predict_2022_hispanicwhite) 
```

### 2023-04-07 products/manuscript/Manuscript.qmd
Submit Part 4

---
title: "MADA 8060E Project"
subtitle: "Trends in New HIV Diagnoses in the US, 2008-2022"
author: Nicole Luisi
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

@@

The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years. Previous research by [Chapin-Bardales et al. (2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013. This analysis will focus on new HIV diagnoses between 2008-2022; subgroups will include region, age group, gender, and race/ethnicity. [@Chapin2017]

@@

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

@@

* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) [@Sullivan2020]

@@

* Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
* Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
* Average annual percent change (AAPC) for incidence (calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation))

## Questions/Hypotheses to be addressed
This analysis will evaluate national and regional trends in HIV new diagnoses between 2008-2020 by sex, age, and race/ethnicity.

@@

## Data aquisition
Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in `data/raw_data/AIDSVu` sub-folders. The new diagnosis files consist of one xlsx file per year for each level (national, regional, etc.).

@@

@@

>>>>

---
title: "MADA 8060E Project"
subtitle: "Trends in New HIV Diagnoses in the US, 2008-2020"
author: Nicole Luisi
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

@@

The purpose of this analysis is to analyze US trends in new HIV diagnoses over the past 10 years (2008-2020) by region, age group, gender, and race/ethnicity. Joinpoint regression was used to identify significant trends and annual percent changes in new diagnoses rates, and trends for rate ratios were also used to assess disparities by race/ethnicity over time. Google Trend data from 2016-2020 for search term "HIV" was used to predict not yet released Black-White and Hispanic-White rate ratios for 2022.

@@

## General Background Information

Previous research by [Chapin-Bardales et al. (2017)](https://doi.org/10.1016/j.annepidem.2017.04.002) evaluated trends in AIDS diagnosis rates by race/ethnicity between 1984-2013.

[@Chapin2017] [@Sullivan2020] [@SULLIVAN2018833] [@FIELDS2021] [@BOSH2021] [@MCCREE2019] [@YOUNG2018]

@@

* HIV New Diagnoses: [AIDSVu](https://aidsvu.org/) 
* Google Trend data for search term "HIV": [Google Trends](https://trends.google.com/)

[@Sullivan2020]

@@

* Rate (per 100,000) of new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
* Rate ratios for new HIV diagnoses by age, gender, and race/ethnicity (national, regional)
* Average annual percent change (AAPC) for incidence (calculated using [Joinpoint Regression Software (Version 4.9.1.0)](https://surveillance.cancer.gov/help/joinpoint/tech-help/citation))
* Predictions for Black-White and Hispanic-White rate ratios by state, using Google Trend search data for the term "HIV"

## Questions/Hypotheses to be addressed
This analysis will evaluate national and regional trends in HIV new diagnoses between 2008-2020 by sex, age, and race/ethnicity. Black-White and Hispanic-White rate ratios will also be predicted for 2022 based on Google Trend results for search term "HIV".

@@

## Data aquisition
Data files were downloaded from [AIDSVu](https://aidsvu.org/) and placed in `data/raw_data/AIDSVu` sub-folders. The new diagnosis files consist of one xlsx file per year for each level (national, regional, state, etc.). For [Google Trends](https://trends.google.com/), annual estimates for search term "HIV" were downloaded for each state.

@@

Google Trend data for search term "HIV" was downloaded for each state and year between 2016-2020. These data were used in linear models to predict not yet released 2022 Black-White and Hispanic-White rate ratios based on Google Trend data for 2022.

@@

@fig-bwrrpredict22 shows state-level predictions for Black-White Rate Ratios in 2022 based on Google Trend data for Search Term "HIV".

```{r}
#| label: fig-bwrrpredict22
#| fig-cap: "Black-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
#| echo: FALSE
knitr::include_graphics(here("results","Predict2022_BWratio.png"))
```

@fig-hwrrpredict22 shows state-level predictions for Hispanic-White Rate Ratios in 2022 based on Google Trend data for Search Term "HIV".

```{r}
#| label: fig-hwrrpredict22
#| fig-cap: "Hispanic-White Rate Ratio Prediction for 2022 based on Google Trend for HIV Search Term"
#| echo: FALSE
knitr::include_graphics(here("results","Predict2022_HWratio.png"))
```
